---
title: "Project_eac238"
author: "Eric Cartaya"
netid: "eac238"
date: "5/1/2020"
output: word_document
---

# A Look At American Politics Over Time
In the modern American political landscape there are two dominant parties, the Democrats and the Republicans, with which you should be familiar with already. There are however many smaller parties and independent candidates that run every election on different platforms, typically to less desirable results. I decided I wanted to look at how these parties have changed over time from two perspectives: how successful they are in presidential elections and how many people identify against the two major parties. 

The first dataset I found after only a couple of hours of looking from the MIT Election Lab, which gave me a link to a Harvard database for some reson. This gave me information of the popular vote in presidential elections over time and was tidied and scraped with relative ease. The other dataset was a massive pain and took me nearly a month (I found it on the 1st of May) to find it. I nearly thought about abandonning the project completely on multiple occacions, but I did not. It also took considerably more effore to scrape it, but was eventually molded into a dataset that showed me the results of polls on people's political identifications over time. 

## The Data and The Analysis

First I will go over the data and results, then I will cover the actual analysis and results.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytext)
library(ggplot2)
library(httr)
library(XML)
library(lubridate)
```

```{r functionSetup}
yeartodate <- function(year)# A function to get the accurate number of days since 1970 to convert into the date format.
{
  day1970 <- ((year-1970)*365) + floor((year-1970)/4)# The math for the conversion.
  day1970 <- day1970 + 320# I chose November 15th as that is the last possible day after the presidential elections taking everything into account, but I got the date wrong so November 16th will have to suffice.
}
```

The first dataset (the MIT Election Lab dataset), had to be changed to relect the national counts for each party above the state counts, but I was able to get a count per election per party. Some parties are divided by states which might skew the numbers, as well as some parties having misspelled names such as the "Communist Party USE". All the major parties are spelled right in all elections from what I can tell with only the Libertarians changing their name halfway as the only major change so I didn't bother correcting anything. 

Down below we have all of the results per election per party, with the results from all elections being graphed. I couldn't add color to the graph as the colors take up the whole graph. A full graph will be in the presentation though. 

```{r presidentialElection_a}
partyPresidentTable <- read.table("https://dataverse.harvard.edu/api/access/datafile/:persistentId?persistentId=doi:10.7910/DVN/42MVDX/MFU99O", sep="\t", header=TRUE)# Using the MIT Election Lab's data (which is held by Harvard for some reason) I got the data for every popular vote for every presidential election since 1976.
partyPresidentTable <- partyPresidentTable %>% mutate(date = as_date(yeartodate(year))) %>% select(-year) %>% group_by(date, party) %>% summarise(candidatevotes = sum(candidatevotes), totalvotes = sum(totalvotes))# Changes the dataset to give the national popular vote by year and by party, and to have an actual date associated with the elections.
print(partyPresidentTable)# Printing the original results.
ggplot(data = partyPresidentTable, aes(x = date, y = candidatevotes)) + geom_line(aes(x = date, y = candidatevotes, fill = party)) + scale_y_continuous(name="Votes", limits=c(100000, 70000000)) + labs(title ="Votes Each Election", x = "Year", y = "Votes")# A much nicer table will be featured later that has lables and properly displays the 3rd party results.
```

This contained all of the total values for each party and ranks them by how many votes they've gotten, which did reveal some issues. Firstly, there is a blank space, which is for blank and spoiled ballots, it is not a mistake and is completely intentional. I believe it was left blank to make sure they wouldn't get confused for the "Spoiled Ballot Party". Secondly two of the largest parties are the Independent Party and the Reform party, which will be explained later. In 1992 both hit their record highs and both ran the same candidate, Ross Perot, who was a big part of the reason that I did this topic in the first place.

```{r presidentialElection_b}
partyNames <- partyPresidentTable %>% select(party, candidatevotes) %>% group_by(party) %>% summarise(partyvotes = sum(candidatevotes)) %>% arrange(desc(partyvotes))# Gets a list of the political parties and their total vote counts. 
print(partyNames)
```

```{r predidentialElection_c}
consolidatedPresidentTableDem <- partyPresidentTable %>% filter(party == "democrat") %>% rename(demvotes = candidatevotes) %>% select(-party)# Gets the number of votes that the Dems got in each election.
consolidatedPresidentTableRep <- partyPresidentTable %>% filter(party == "republican") %>% rename(repvotes = candidatevotes) %>% select(-party)# Same but for the GOP.
consolidatedPresidentTable <- full_join(consolidatedPresidentTableDem, consolidatedPresidentTableRep, by = "date") %>% select(-totalvotes.x) %>% rename(totalvotes = totalvotes.y) %>% mutate(thirdpartyvotes = totalvotes - demvotes - repvotes)# Combines the Dems and GOP so that we can use the total votes cast to see the number of votes that went to third party and independent candidates. Gather was also used so that the votes were all in one row.
print(consolidatedPresidentTable)# Printing the edited results.
ggplot(data = consolidatedPresidentTable, aes(x = date)) + geom_line(aes(y = repvotes, color = "darkred")) + geom_line(aes(y = thirdpartyvotes, color = "green")) + geom_line(aes(y = demvotes, color = "steelblue")) + labs(title ="Votes Each Election", x = "Year", y = "Votes")# Printed a graph with the election resuslts for the combine 3rd parties. Note the spikes being Ross Perot in 1992 and basically anyone in the 2016 election. 
write.csv(consolidatedPresidentTable,"C:\\Users\\Eric\\Downloads\\president.csv", row.names = TRUE)# Ejecting data to a CSV.
```

```{r election_model1}
consolidatedPresidentTable2 <- consolidatedPresidentTable %>% select(-totalvotes) %>% gather("party", "votes", demvotes, repvotes, thirdpartyvotes)# This is the building of a model to see how time has affected each of the political parties. This has lead me to the conclusion that political plurality tends to be weak and that the Democrats and Republicans politically dominate the precidency to a total degree (what a suprise).
presm <- lm(votes ~ ., data = consolidatedPresidentTable2)
summary(presm)
```

```{r election_model2}
consolidatedPresidentTable3 <- consolidatedPresidentTable %>% gather("party", "votes", demvotes, repvotes, thirdpartyvotes)# This was much of the same but I decided to keep the total number of votes in the model, creating a monstrosity that I kept just to show off the P-value of 1.
presm2 <- lm(votes ~ ., data = consolidatedPresidentTable3)
summary(presm2)
```

```{r gallupPoll_a}
writ <- GET(url = "https://news.gallup.com/poll/15370/party-affiliation.aspx")# This was to scrape data from the Gallup Polling website, and yes this was the best I could do for a second dataset from a second website.
writXML <- htmlParse(content(writ, as = "text"))# All of this was to get the scraped table into a workable dataset.
writTable <- getNodeSet(writXML, '//*[contains(concat( " ", @class, " " ), concat( " ", "figure-table", " " ))]')
rowGetter <-sapply(writTable, xmlValue)
gallupPol <- rowGetter[1]# All of this is to get the correct table and to remove all of the strings that are unusable.
gallupPol <- str_split(gallupPol, "\\r\\n")
gallup <- as.data.frame(gallupPol)# Converts the data into a dataframe.
gallup <- gallup %>% rename(x = c..In.politics..as.of.today..do.you.consider.yourself.a.Republican..a.Democrat.or.an.independent....) %>% filter(x != "%") %>% filter(x != " ") %>% filter(x != "Republicans") %>% filter(x != "Democrats") %>% filter(x != "Independents") %>% filter(x != "Trend since 2004") %>% filter(x != "In politics, as of today, do you consider yourself a Republican, a Democrat or an independent?") %>% filter(x != "") %>% filter(x != "Gallup")
head(gallup)# Displays initial dataset.
```

```{r gallupPoll_b}
gallupDem <- gallup[seq(4, nrow(gallup), 4), ] %>% as.data.frame()# Gets the correct percentages for each of the parties. 
gallupInd <- gallup[seq(3, nrow(gallup), 4), ] %>% as.data.frame()
gallupRep <- gallup[seq(2, nrow(gallup), 4), ] %>% as.data.frame()
gallupDat <- gallup[seq(1, nrow(gallup), 4), ] %>% as.data.frame()# This one gets the dates.
gallupDem <- gallupDem %>% rename('Democrats %' = '.')# This corrects the names.
gallupInd <- gallupInd %>% rename('Independents %' = '.')
gallupRep <- gallupRep %>% rename('Republicans %' = '.')
gallupDat <- gallupDat %>% rename(Date = '.')
gallup <- bind_cols(gallupDat, gallupRep)# Binds the tables together.
gallup <- bind_cols(gallup, gallupInd)
gallup <- bind_cols(gallup, gallupDem)
gallup <- gallup %>% mutate(`Republicans %` = as.numeric(as.character(`Republicans %`))) %>% mutate(`Independents %` = as.numeric(as.character(`Independents %`))) %>% mutate(`Democrats %` = as.numeric(as.character(`Democrats %`))) %>% mutate(Date = as_date(substr(gallup$Date, 0, str_locate(gallup$Date, "-")[,1]-1))) %>% gather("party", "percent", `Republicans %`, `Independents %`, `Democrats %`)# Fixes the dates and the parties so that all the percentages so that they are in one party.
head(gallup)# Displays the final dataset
ggplot(data = gallup, aes(x = Date, y = percent, color = party)) + geom_line(aes(x = Date, y = percent, color = party)) + scale_y_continuous(name="Percentage", limits=c(0, 50)) + labs(title ="Percentage of Population", x = "Year", y = "Percentage")# Shows tha popularity of these groups since 2007
write.csv(gallup,"C:\\Users\\Eric\\Downloads\\gallup.csv", row.names = TRUE)# Ejecting data to a CSV
```

```{r gallup_model}
gallup <- gallup %>% mutate(parInd = 0) %>% mutate(parRep = 0) %>% mutate(parDem = 0)# Added in indicator values, this ended up being a waste of time as you get the same exact result as with just using the native string indicator in the modeling function.

for (i in 1:(nrow(gallup)/3))# Added in indicators by party using the position that I know they are in.
{
  gallup[[5]][i] = 1
}
for (i in (nrow(gallup)/3+1):(2*nrow(gallup)/3+1))
{
  gallup[[4]][i] = 1
}
for (i in (2*nrow(gallup)/3+1):nrow(gallup))
{
  gallup[[6]][i] = 1
}

pm <- glm(percent~Date+parInd+parRep+parDem, data = gallup)# Modeled the data and discovered that most people are unwilling to declare as Democrats or Republicans (all third parties got lumped in with independent as per the article) and the popularity of independence is increasing greatly with time. 
summary(pm)
```
